\section{Discussion}

\subsection{What the contradictions are---and are not}

Taken together, the results support a single claim: the apparent inter-study contradictions
in aluminum--water hydrolysis are \emph{predominantly methodological}. Naive pooling
overstates predictability by $0.62$--$0.85$ in $R^2$; the physico-chemical regime label
explains $\approx\!2\%$ of between-study yield variance while methodological covariates
explain $\approx\!55\%$; the flagship particle-size contradiction dissolves. The practical
implication is direct: a hydrogen yield reported without its measurement protocol is only
weakly comparable across laboratories, and meta-analyses that pool raw yields inherit
methodological variance as if it were chemistry.

Naming the specific apparent contradictions helps calibrate the claim. For activation
energy, the most striking contrast in the liquid-metal regime is between
\citeauthor{ilyukhina2012}~\citep{ilyukhina2012} ($E_\mathrm{a}\!=\!8.5$\,kJ\,mol$^{-1}$,
Ga--In eutectic, peak-rate proxy) and \citeauthor{dudoladov2016}~\citep{dudoladov2016}
($E_\mathrm{a}\!=\!58.0$\,kJ\,mol$^{-1}$, Ga in KOH, tabulated rate): both are
liquid-metal-activated systems, yet they differ by nearly sevenfold. Our analysis does
not resolve whether this reflects different Ga/In loading, different activation proxies
(peak rate vs.\ initial rate), or different surface geometries; it demonstrates that the
current ``liquid\_metal\_activated'' label is too coarse to resolve this.
For particle size, cross-study comparisons are misleading because studies compare
incompatible size windows: one laboratory's ``fine'' particles are another's standard
size, and different regimes are mixed when averaging across studies. Within each of the
five studies that vary particle size, the effect is consistent and negative (smaller
$\rightarrow$ higher yield; Spearman $\rho\!=\!-0.29$ to $-0.87$ in
\citep{zhang2024,prabu2021,wen2018,porciuncula2012}). The apparent cross-study flip is
thus an artifact of comparing different experimental windows, not a genuine physics
disagreement.

\subsection{Why pooling fails: the cross-study evidence in context}

The scale of the optimism gap ($0.62$--$0.85$) is consistent with what has been found
when grouped cross-validation is applied to other domains with hidden group structure
\citep{bernett2024,john2025}. The closest published analogues for hydrogen generation
are the interlaboratory round-robin comparisons for MgH$_2$ storage \citep{wulf2021} and
biohydrogen fermentation \citep{hoque2018}: in both cases, identical ``conditions'' in
different laboratories produced yields spanning roughly twofold, with the spread traced
primarily to measurement and protocol choices rather than to genuine material differences.
The present result---that temperature control alone explains $33\%$ of between-study
yield variance in aluminum hydrolysis---places this field in the same category. The
leakage-controlled benchmark we release is designed to let future modeling efforts detect
the same failure mode: any model that achieves positive grouped $R^2$ on the provided
corpus under the study-grouped split would represent genuine cross-study generalization
beyond what the current literature supports.

\subsection{Physics is not absent}

``Predominantly methodological'' is not ``purely methodological.'' The regime label is
coarse; finer, particle-size-dependent kinetics---tunable between nano- and
micro-aluminium powders---genuinely shape the reaction \citep{saceleanu2019}, and such
physics operates \emph{within} studies, below the resolution of a five-level label and
partly absorbed into the methodological covariates with which it co-varies. Incremental
regime variance ($+0.15$ after controlling for method) is nonzero, even if not robust.
Our attribution is also observational: because methodology is largely a study-level
attribute, the temperature-control association could be partly confounded with other
unmeasured study-level choices (principal investigator laboratory practice, instrument
calibration, student experience). We state the finding associationally and quantify,
rather than assert, its robustness.

\subsection{A minimum-information reporting standard}

The remedy is standardization. Table~\ref{tab:mireport} lists the minimum information
needed to interpret a hydrogen yield across laboratories. The ordering reflects the
between-study variance contribution each field carried in our analysis: temperature
control is listed first because it was the single largest source ($R^2\!=\!0.33$);
measurement method and water type follow because they contribute smaller but nonzero
shares. The remaining fields---alkali type and concentration, particle size, activation
route, yield basis, rate definition---carried detectable but weaker or less-covered
signal here, yet they are practically essential for reproducibility. Provenance (DOI,
condition count, replicate count) is necessary to build cumulative, poolable evidence.

This template is modelled on reporting standards that successfully improved poolability
in related fields: round-robin-derived checklists for MgH$_2$ \citep{wulf2021} and
biohydrogen \citep{hoque2018} both show that standardization shrinks interlaboratory
spread rather than eliminating it. We expect the same here: full compliance would not
remove between-study variance, but it would expose the remaining variance as physics
rather than protocol.

\begin{table}[t]\centering\small
\caption{Proposed minimum-information reporting standard for aluminum--water hydrolysis
hydrogen-yield studies. Items are ordered by the between-study yield variance each
carried here (Section~\ref{sec:variance_decomp}); temperature control is the largest
single source ($R^2\!=\!0.33$).}
\label{tab:mireport}
\begin{tabular}{lll}
\toprule
Field & Report & Between-study $R^2$ here \\
\midrule
Temperature control  & method (isothermal bath / self-heating / uncontrolled) + setpoint & 0.33 \\
Measurement method   & water displacement / pressure / GC / mass-flow & 0.07 \\
Water type \& purity & DI / tap / alkaline / saline, with conductivity or grade & 0.04 \\
Quality / control    & replication count; value reported vs.\ derived & 0.04 \\
Reaction vessel      & open or closed; insulated? & 0.02 \\
Alkali               & type (NaOH/KOH/none) + concentration (mol\,L$^{-1}$) & --- \\
Particle size        & $D_{50}$ ($\mu$m) + sizing method & --- \\
Activation route     & pure / melt-alloyed / ball-milled / liquid-metal / waste & --- \\
Yield basis          & theoretical reference (mL\,g$^{-1}$); \% vs.\ volumetric & --- \\
Rate definition      & initial / maximum / average; $t_{50}$/$t_{80}$ if rate is claimed & --- \\
Provenance           & study DOI, number of conditions, replicate count & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Limitations}
\label{sec:limitations}

\paragraph{Sample size and power.}
The analysis rests on $n\!\approx\!31$ studies ($3$--$9$ per regime). Null results
(no regime moderation of yield, H2; no regime structure in $E_\mathrm{a}$, H3;
heterogeneous within-study predictability, T\_A) are reported as ``no detectable effect
at this sample size,'' not as proof of absence. The between-study variance decomposition
(T\_B) has cluster-bootstrap confidence intervals that are wide at this $n$; the lower
bound on method-joint $R^2$ ($0.36$) excludes zero but the upper bound ($0.999$) is
uninformative. Two regimes (\texttt{waste\_al}, \texttt{liquid\_metal\_activated}) have
fewer than $40$ yield rows each; effect estimates for these classes are exploratory.

\paragraph{Associational attribution.}
All findings are associational, not causal. Because temperature-control method is largely a
study-level attribute, its $33\%$ variance share could partly reflect unmeasured
study-level differences (equipment, practice, calibration) that co-vary with protocol
choice. We quantify the incremental contribution (method beyond regime: $+0.47$; regime
beyond method: $+0.15$) but do not attribute a causal mechanism.

\paragraph{Single material system and open-access bias.}
The dataset covers only aluminum--water hydrolysis; the leakage finding and the
methodological-heterogeneity finding may not transfer to other reaction systems, though
analogues exist in MgH$_2$ and biohydrogen \citep{wulf2021,hoque2018}. Open-access
retrieval may bias the sample toward English-language, higher-impact, more-systematic
studies; we mitigate this with a quality tier and report the access constraint here.

\paragraph{Unvalidated reporting standard.}
The proposed minimum-information checklist (Table~\ref{tab:mireport}) has not been
prospectively validated: we do not yet know how much of the residual between-study
variance it would remove in practice. It is a hypothesis about what would help, grounded
in our variance decomposition.

\paragraph{Single-author and AI-assisted workflow.}
All data extraction, curation, analysis, and writing were performed by a single author
with AI assistance (see Declaration of generative AI). Single-author extraction carries
no inter-rater reliability guarantee beyond the internal double-extraction check
($52/52$ exact, covering $16\%$ of yield rows, $52/315$); the remaining $263$ rows
have no independent extractor check, leaving systematic extraction errors undetectable
by this protocol. Because both extraction passes may have relied on the same AI tool,
the $52/52$ agreement is a self-consistency check rather than an independent validation
against AI-introduced errors---the primary protection is the author's manual review and
verification of every extracted value. Readers should treat the results as a first
systematic synthesis requiring external validation rather than as an established consensus.
