\section{Introduction}
The reaction of aluminum with water, $2\,\mathrm{Al}+3\,\mathrm{H_2O}\rightarrow
\mathrm{Al_2O_3}+3\,\mathrm{H_2}$ (or via hydroxides), is an attractive route to
hydrogen on demand: it is triggered by water alone, stores energy at high density, and
needs no high-pressure H$_2$ infrastructure. Two decades of experiments have mapped how
temperature, alkalinity, particle size, alloying/activation, and water chemistry control
the yield and rate. Yet the resulting literature is \emph{internally contradictory}.
Reported activation energies span $3.5$--$102.6$\,kJ\,mol$^{-1}$; the effect of particle
size is reported as both beneficial and detrimental; the rate--yield trade-off is
inconsistent; and studies disagree on which parameter dominates. These contradictions are
usually attributed, in passing, to ``different conditions,'' but the attribution has never
been tested.

This is the unresolved question we address: \emph{are the apparent inter-study
contradictions genuine physico-chemical regime differences, or artifacts of isolated,
non-standardized experimental conditions?} The distinction matters because it determines
whether the literature can be pooled to build predictive design models at all. Prior
machine-learning work on aluminum--water hydrolysis is, to our knowledge, limited to a
single-study, in-sample neural network reporting near-perfect prediction
($R^2\!\approx\!0.998$; \todocite{das2023fuel}). Such a number is exactly the
generalization illusion that a cross-study analysis must guard against: a model that
interpolates one laboratory's data says nothing about whether the literature agrees.

We take the opposite stance. We unify the scattered literature into a single
provenance-tracked dataset and ask, with leakage-controlled evaluation, how far one study
transfers to another, and \emph{what} drives the differences when it does not. We
pre-registered the analyses and their success criteria before computing them, and we
report nulls transparently as ``no detectable effect at this sample size'' rather than as
evidence of absence.

\paragraph{Contributions.}
\begin{itemize}\itemsep2pt
  \item A curated, provenance-tracked, quality-tiered \textbf{open dataset} of 315
  condition-level hydrogen-yield rows from 31 studies, with a separate 76-row kinetic
  table (activation energies and rates), released under CC-BY.
  \item A \textbf{leakage-controlled evaluation} showing that random-split predictability
  ($R^2\!\approx\!0.55$) collapses under study-grouped cross-validation ($R^2<0$; optimism
  gap $0.62$--$0.85$)---the literature does not pool naively.
  \item The central finding: a study-level variance decomposition in which
  \textbf{methodological covariates explain $\approx\!55\%$ of the between-study yield
  variance versus $\approx\!2\%$ for the physico-chemical regime}, with the
  temperature-control method alone the largest single source ($\approx\!33\%$); the effect
  survives controlling for regime.
  \item Resolution of the \textbf{flagship particle-size contradiction} as a cross-study
  integration artifact (within-study effect negative in $5/5$ studies).
  \item A proposed \textbf{minimum-information reporting standard} for aluminum-hydrolysis
  studies, to make future data poolable.
\end{itemize}
