\section{Data and methods}
\subsection{Provenance-tracked dataset}
We screened the aluminum--water hydrolysis literature and extracted, from 31 full-text
studies in hand, every distinct experimental condition reporting a hydrogen outcome. Each
row records the target (H$_2$ yield, \% of the $1244$\,mL\,g$^{-1}$ pure-Al theoretical),
physical parameters (temperature, alkali concentration, particle size, activator ratio,
elemental composition), categorical descriptors, full provenance (study DOI, table/figure
source, extractor, extraction method), and quality/methodology metadata. The grouping key
is the study DOI. Absent quantities are encoded as $0$ and unreported quantities as
missing, never conflated. Each study is assigned a physico-chemical regime label
\texttt{system\_class} $\in$ \{\texttt{pure\_al\_alkali}, \texttt{al\_alloy},
\texttt{mechanically\_activated}, \texttt{liquid\_metal\_activated}, \texttt{waste\_al}\},
verified against the source materials/methods. The final yield dataset has 315 rows from 31
studies (Fig.~\ref{fig:dataset}); a \emph{separate} 76-row kinetic table holds activation
energies and rates (reported or fit from rate-vs-temperature data), with their own
provenance and fit-quality fields, and is never merged into the yield table.

Two independent re-extraction passes provide quality control: a double extraction of
$\approx\!16\%$ of yield rows showed $52/52$ exact value agreement, and a classification
audit re-derived \texttt{system\_class} for all 31 studies from source; digitized kinetic
points were re-read independently ($\approx\!1.5\%$ point error). Studies whose Arrhenius
fits failed a quality gate ($R^2\!\ge\!0.90$, $\ge\!3$ temperatures) were dropped rather
than forced.

\subsection{Leakage-controlled evaluation}
For each of six comparative regressors (ElasticNet, $k$NN, random forest, XGBoost, LightGBM,
Gaussian process) we estimate the held-out coefficient of determination under two protocols:
ordinary $k$-fold (records split at random) and \emph{study-grouped} $k$-fold (all records
of a study held out together). The \emph{optimism gap}, $R^2_{\text{random}}-
R^2_{\text{grouped}}$, measures how much random splitting overstates cross-study
generalization. Tree models retain missing values; linear/kernel models use median
imputation and one-hot encoding.

\subsection{Between-study variance decomposition (primary analysis)}
To ask \emph{what} accounts for the disagreement, we aggregate to the study level (one row per
study: mean yield and the study's modal covariate values) and decompose the between-study
variance of mean yield. For a categorical predictor $C$ we report the variance share
$R^2_C = 1 - \mathrm{SS_{within}}(C)/\mathrm{SS_{total}}$, comparing the physical-regime
label against methodological covariates (\texttt{temperature\_control},
\texttt{measurement\_method}, \texttt{water\_type}, \texttt{quality\_tier},
\texttt{vessel\_type}). We test whether the methodological signal is genuine---rather than a
re-encoding of study identity or a confound with regime---by (i) auditing the study-level
coverage of each covariate level, (ii) cross-tabulating against \texttt{system\_class}, and
(iii) reporting the \emph{incremental} variance explained by the temperature-control method
after controlling for regime. All headline numbers carry cluster-bootstrap confidence
intervals (resampling studies); contrasts use Holm correction. We treat the kinetic
$E_\mathrm{a}$ analysis and the within-study tests analogously.

\subsection{Discipline}
Hypotheses and success thresholds were frozen before computation. Language on observational
findings is associational, not causal. We report nulls as ``no detectable effect at
$n\!\approx\!31$ studies / $3$--$9$ per regime.'' Machine-learning attributions
(SHAP/importances) are used descriptively only. All numbers in this paper trace to a
committed analysis artifact, and the full pipeline is released.
