\section{Related work}
\label{sec:related}

\paragraph{Pure alkali and early rate studies.}
The majority of the literature covers aluminum activated by dissolved NaOH or KOH in
aqueous solution---the simplest and most studied regime. Early kinetic work by
\citeauthor{porciuncula2012}~\citep{porciuncula2012} established systematic temperature and
concentration factorial data for NaOH–foil experiments that remain the largest single-study
block in the open literature. Subsequent studies report yields and rates under broadly
similar conditions but incompatible protocols
\citep{wen2018,wanghq2017,yavor2013,trowell2022,deng2010,prabu2021,rin2021,zhang2024}.
Activation energies for this regime cluster around $50$\,kJ\,mol$^{-1}$ in well-designed
studies but span $37$--$65$\,kJ\,mol$^{-1}$ across the literature.

\paragraph{Mechanically activated and alloy systems.}
Ball milling with low-melting additives (Sn, Bi, In) lowers the activation barrier by
disrupting the native oxide layer and is the subject of a growing sub-literature
\citep{xiao2018,guan2019,preez2018,chen2020,xiao2020,davies2022mat}. Al-alloy systems
(pre-alloyed Sn, Ga, In) operate through a related but distinct dissolution mechanism
\citep{liuyh2017,qiao2019,meng2022}. Both groups exhibit higher reported yields at
moderate temperatures than the pure-alkali regime, but direct comparisons are complicated
by incompatible particle-size and additive specifications.

\paragraph{Liquid-metal activated systems.}
Gallium-based eutectic infiltration produces a highly active surface with dramatically
reduced apparent activation energies: \citeauthor{ilyukhina2012}~\citep{ilyukhina2012}
report $E_\mathrm{a}\!=\!8.5$\,kJ\,mol$^{-1}$ for a Ga--In eutectic system, while
\citeauthor{dudoladov2016}~\citep{dudoladov2016} report $E_\mathrm{a}\!=\!58.0$\,kJ\,mol$^{-1}$
for a different composition, and \citeauthor{jayaraman2015}~\citep{jayaraman2015} and
\citeauthor{lu2017}~\citep{lu2017} report intermediate values. The $8.5$--$58.0$\,kJ\,mol$^{-1}$
within-regime spread is the headline apparent contradiction for this class and reflects
varying Ga loading, alloy purity, and activation pre-treatment
\citep{fischman2020}. Minimum-reporting-standard violations mean that many results are
not directly comparable even within this regime.

\paragraph{Waste and recycled aluminum.}
Studies using waste or scrap aluminum have grown in number, motivated by the
simultaneous valorization of waste streams and hydrogen generation
\citep{gupta2025,ho2016,knoks2025,mezulis2023,martinezv2026,buryakov2023met,urbonav2024,fadhilah2023}.
Reported yields vary widely, partly because the composition of waste aluminum is
heterogeneous and poorly characterized, and partly because activation pre-treatments
(sorting, cleaning, mechanical grinding) differ across laboratories. This regime has the
highest fraction of unreported experimental parameters in our corpus.

\paragraph{Reviews: cataloguing without resolving.}
Narrative reviews of the field
\citep{musicco2025,dupreez2021review} catalogue the apparent contradictions and
emphasize the breadth of strategies, but do not pool the data, do not control for
methodological heterogeneity, and do not test whether the contradictions are regime-driven
or protocol-driven. The gap between the experimental volume and the absence of a pooled,
quantitative resolution is the direct motivation for the present work.

\paragraph{The generalization illusion in single-study ML models.}
Data-driven modeling of this reaction is nascent. The closest prior work is a single-study
artificial neural network reporting a near-perfect in-sample fit within one laboratory's
data \citep{das2023fuel}. In-sample performance on one study's parameter grid is
uninformative about cross-study transfer; our leakage analysis makes this gap explicit.
This failure mode is not unique to aluminum hydrolysis: it has been documented in materials
screening \citep{bernett2024,john2025}, chemometrics \citep{pomerantsev2021}, and catalysis
data science \citep{suvarna2024}, and is increasingly recognized as a reproducibility risk
in AI-driven materials and energy research \citep{bozalginesta2025,coelho2022,xue2024,noble2022}.

\paragraph{Leakage-controlled evaluation in materials science.}
Grouped or blocked cross-validation---where all records from one group are withheld
together---is the established defense against optimistic within-group estimates
\citep{bernett2024,john2025}. The approach closely parallels leave-one-group-out designs
in meta-analysis and the block-bootstrap for correlated data. We apply it here to
aluminum hydrolysis for the first time; beyond demonstrating leakage, we use the
between-study variance structure to \emph{attribute} the disagreement quantitatively.

\paragraph{Methodology as a confound, and reporting standards.}
Individual studies already note that protocol choices shift measured values substantially:
vessel insulation and temperature control \citep{urbonavicius2023}, water type, and
surface pre-treatment \citep{wen2018} all appear as study-specific idiosyncrasies that
prevent pooling. Minimum-information reporting standards have improved poolability in
related fields---round-robin interlaboratory comparisons for MgH$_2$
\citep{wulf2021} and biohydrogen \citep{hoque2018}---demonstrating that standardized
protocols can turn an apparently contradictory literature into a comparable one. We adapt
and propose such a template for aluminum hydrolysis.
