% International Journal of Hydrogen Energy submission — Elsevier elsarticle class.
% Compiles on Overleaf / TeX Live (CI). Local fallback: if elsarticle is unavailable, the previous
% `article`-class preamble is preserved in git history (commit before the elsarticle conversion);
% the section files and references are class-independent, so only this preamble would need swapping.
\documentclass[preprint,12pt]{elsarticle}

\usepackage[T1]{fontenc}   % correct rendering of accented characters (Çankaya, Türkiye)
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}   % \blacksquare etc. used in figure captions
\usepackage{xcolor}
\usepackage[colorlinks=true,allcolors=blue]{hyperref}
% elsarticle loads natbib internally — do NOT load natbib again here.
\graphicspath{{figures/}}
\newcommand{\todocite}[1]{\textcolor{red}{[cite: #1]}}  % placeholder macro (currently unused)
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}

\journal{Energy and AI}

\begin{document}

\begin{frontmatter}

\title{Apparent Contradictions in the Aluminum--Water Hydrolysis Literature
are Predominantly Methodological: a Provenance-Tracked, Leakage-Controlled Analysis}

\author[a]{Dogukan Unal\corref{cor}}
\ead{dunal@ipec.com.tr}
\affiliation[a]{organization={IPEC, Industrial Project Engineering Consulting},
                city={Çankaya, Ankara}, postcode={06550}, country={Türkiye}}
\cortext[cor]{Corresponding author. ORCID: 0009-0006-5102-8013
              (\url{https://orcid.org/0009-0006-5102-8013}).}
%% Single author: Dogukan Unal.

\begin{abstract}
\noindent Two decades of aluminum--water hydrolysis experiments have produced a literature
that contradicts itself: reported activation energies span $3.5$--$102.6$\,kJ\,mol$^{-1}$
and the direction of the particle-size effect changes sign across studies. Whether these
contradictions reflect genuine physico-chemical regimes or artifacts of non-standardized
protocols is the question we address with a leakage-controlled machine-learning evaluation
framework. We curate a provenance-tracked, quality-tiered open dataset of 315 condition-level
hydrogen-yield rows from 31 independent studies (plus a separate 76-row kinetic table), with
pre-registered hypotheses and transparent null reporting. Under study-grouped cross-validation,
six comparative machine-learning models collapse from $R^2\!\approx\!0.55$ (random splits)
to $R^2<0$ (optimism gap $0.62$--$0.85$)---the literature does not pool naively. A
between-study variance decomposition attributes $\approx\!55\%$ of the between-study
yield variance to \emph{methodological} covariates versus $\approx\!2\%$ for the
physico-chemical regime label, with temperature-control method alone the largest single
source ($\approx\!33\%$). A complementary mixed-effects meta-regression finds no significant
regime moderation ($p=0.32$--$0.45$, $0/8$ Holm-corrected interactions). The flagship
particle-size ``contradiction'' dissolves: within every study that varies particle size
($5/5$), the within-study effect is consistently negative, identifying the cross-study
apparent flip as an integration artifact. We release the curated dataset, the reproducible
analysis pipeline, and a minimum-information reporting standard to make future experiments
poolable across laboratories.
\end{abstract}

\begin{keyword}
aluminum--water hydrolysis \sep hydrogen production \sep data leakage \sep
meta-analysis \sep reproducibility \sep open dataset
\end{keyword}

\end{frontmatter}

%% Highlights (3--5 bullets, each $\le$85 chars). For Elsevier upload these are also provided as a
%% separate file: paper/submission/highlights.txt. Shown here so the preprint PDF carries them.
\section*{Highlights}
\begin{itemize}\setlength{\itemsep}{0pt}
  \item Leakage-controlled ML evaluation: $R^2\!\approx\!0.55$ (random); $R^2\!<\!0$ (grouped); gap 0.62--0.85.
  \item Open dataset: 315 yield rows + 76 kinetic rows from 31 studies, CC-BY.
  \item Methodology explains $\approx$55\% of between-study yield variance vs $\approx$2\% for regime.
  \item Temperature-control method alone is the single largest variance source ($\approx$33\%).
  \item Particle-size ``contradiction'' is an integration artifact: 5/5 within-study effects negative.
\end{itemize}

\input{sections/introduction}
\input{sections/related_work}
\input{sections/method}
\input{sections/results}
\input{sections/discussion}
\input{sections/conclusion}

\bibliographystyle{elsarticle-num}
\bibliography{references}

\input{sections/supplementary}

\end{document}
