\section{Conclusion}
We unified the scattered aluminum--water hydrolysis literature into a provenance-tracked,
quality-tiered open dataset and used leakage-controlled, pre-registered analysis to test
whether its long-documented contradictions are physical or methodological. The data answer
clearly, if associationally: under study-grouped evaluation naive pooling fails, the
physico-chemical regime explains $\approx\!2\%$ (bootstrap 95\,\% CI $[0.01,\,0.37]$) of the between-study yield variance while
methodological covariates explain $\approx\!72\%$ (hydrogen measurement apparatus alone $\approx\!43\%$;
temperature control $\approx\!8\%$),
and the flagship particle-size contradiction is consistent within studies. The contradictions
are predominantly an artifact of non-standardized measurement, not genuine regime
disagreement. Beyond the finding, the contribution is durable infrastructure: an open
dataset, a reproducible leakage-controlled benchmark, and a proposed minimum-information
reporting checklist that, if prospectively adopted and validated, could reduce methodological
variance and improve poolability.

\section*{Data and code availability}
The curated dataset (315 condition-level yield rows and the separate 76-row kinetic table), the
analysis pipeline, and the figure-generation scripts are released openly. The curated dataset is
openly available at Zenodo, \url{https://doi.org/10.5281/zenodo.20751918} (CC-BY-4.0). The Zenodo
deposit comprises the curated dataset, analysis scripts, and figure-generation code only; draft
manuscripts, command history, and other internal workflow artifacts are excluded from the public
release. A dataset card documenting the schema, provenance, quality tiers, and scope accompanies
the deposit. Analysis code is available from the corresponding author on reasonable request.

\section*{Acknowledgements}
This research was not supported by external funding.

\section*{Declaration of competing interest}
The author declares no competing financial or personal interests.

\section*{CRediT authorship contribution statement}
\textbf{Dogukan Unal:} Conceptualization, Methodology, Software, Formal analysis, Data curation,
Writing -- original draft, Writing -- review \& editing, Visualization. Single-author contribution encompasses all CRediT roles.

\section*{Declaration of generative AI and AI-assisted technologies}
% AUTHOR: confirm accuracy & completeness
During the preparation of this work the author used generative AI agents (Anthropic's Claude, via
the Claude Code environment) for literature triage, assistance with data extraction from source
papers, statistical analysis and analysis code, figure generation, and drafting of the manuscript.
All AI-assisted steps were carried out under the author's direction. The author reviewed and verified
all extracted data, analyses, results, and claims, edited the AI output as needed, and takes full
responsibility for the content of this publication. The AI tools are not listed as authors and made
no independent scientific judgements.
