\section{Introduction}

The reaction of aluminum with water,
$2\,\mathrm{Al}+3\,\mathrm{H_2O}\rightarrow\mathrm{Al_2O_3}+3\,\mathrm{H_2}$
(or via hydroxides), is an attractive route to hydrogen on demand: it is triggered by
water alone, stores energy at high density, and needs no high-pressure H$_2$
infrastructure. Two decades of experiments have mapped how temperature, alkalinity,
particle size, alloying or activation method, and water chemistry control the yield and
rate. Yet the resulting literature is internally contradictory. Reported apparent
activation energies span $3.5$--$102.6$\,kJ\,mol$^{-1}$; within the single
liquid-metal-activated regime this spread is still $8.5$--$58.0$\,kJ\,mol$^{-1}$
\citep{ilyukhina2012,dudoladov2016}. This range is too large to be explained by
measurement noise alone and, if genuine, would imply fundamentally different rate-limiting
mechanisms even within nominally identical systems. The direction of the particle-size effect changes
sign when studies are compared across the literature, yet within every study that varies
particle size the effect is consistently negative
\citep{zhang2024,prabu2021,wen2018,porciuncula2012}. The rate--yield trade-off
is inconsistent; and studies disagree on which parameter dominates. These contradictions
are usually attributed, in passing, to ``different conditions,'' but the attribution has
never been tested quantitatively.

This is the unresolved question we address: \emph{are the apparent inter-study
contradictions genuine physico-chemical regime differences, or artifacts of isolated,
non-standardized experimental conditions (e.g., temperature-control method, measurement
protocol, water type)?} The distinction matters because it determines
whether the literature can be pooled to build predictive design models. Prior
machine-learning work on aluminum--water hydrolysis is, to our knowledge, limited to a
single-study, in-sample neural network reporting near-perfect prediction
\citep{das2023fuel}. Such a near-perfect in-sample fit is exactly the generalization
illusion that a cross-study analysis must guard against: a model that interpolates one
laboratory's parameter grid says nothing about whether the literature agrees.

We take the opposite stance. We unify the scattered literature into a single
provenance-tracked dataset, apply leakage-controlled evaluation to test how far one study
transfers to another, and use between-study variance decomposition to identify
\emph{what} accounts for the differences when transfer fails. We pre-registered the
analyses and their success criteria before any analysis code was executed and report nulls as ``no
detectable effect at this sample size'' rather than as evidence of absence.

\paragraph{Contributions.}
\begin{itemize}\itemsep2pt
  \item A curated, provenance-tracked, quality-tiered \textbf{open dataset} of 315
  condition-level hydrogen-yield rows from 31 independent studies, with a separate
  76-row kinetic table (activation energies and rates), to be released under CC-BY upon publication; this
  dataset serves as a reproducible \textbf{ML benchmark} for cross-study generalization
  in aluminum--water hydrolysis.
  \item A \textbf{leakage-controlled evaluation} showing that random-split predictability
  ($R^2\!\approx\!0.55$) collapses under study-grouped cross-validation ($R^2<0$; optimism
  gap $0.62$--$0.85$): the literature does not pool naively. A \textbf{negative-control
  robustness check} confirms that the apparent within-class signal is porciuncula-weighted
  (row-weighted LOSO $+0.83$) and not a genuine per-study replication signal (study-macro
  LOSO $-0.31$).
  \item The primary quantitative finding: a study-level variance decomposition in which
  \textbf{methodological covariates explain $\approx\!55\%$ of the between-study yield
  variance versus $\approx\!2\%$ for the physico-chemical regime}, with the
  temperature-control method the largest single source ($\approx\!33\%$); the effect
  survives controlling for regime (incremental $R^2\!=\!+0.47$, bootstrap CI excludes 0).
  \item Resolution of the \textbf{flagship particle-size contradiction} as a cross-study
  integration artifact: within-study effects are consistently negative in all $5/5$
  studies that vary particle size ($\rho\!=\!-0.29$ to $-0.87$).
  \item Transparent reporting of \textbf{null results}: mixed-effects regime moderation
  (H2, $p\!=\!0.32$--$0.45$) and Arrhenius activation-energy regime structure (H3,
  $\eta^2\!=\!0.36$, $p\!=\!0.13$) are both not supported at the pre-registered bar;
  we frame these as power-limited findings at $n\!\approx\!31$ studies.
  \item A proposed \textbf{minimum-information reporting standard} for aluminum-hydrolysis
  studies to make future data poolable across laboratories (not yet prospectively
  validated; see Section~\ref{sec:limitations}).
\end{itemize}
